A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-02-20 Papers/Torque_Paper/content.tex
minor updates
- reducing text
- updating references

As a consequence of the relatively strong model assumptions at the basis of BEM theory, the results from BEM codes can be subject to significant inaccuracies or uncertainties. For example, the effect of sheared inflow \cite{Madsen2012} is not naturally accounted for in the theory and needs to be incorporated via correction terms. Other major model uncertainties in BEM models are for example the time constant in dynamic stall models, the wake correction factor, the tip loss model parameter, and the lift- and drag-polars used to compute local aerodynamic forces. Especially for increasing turbine sizes, these model parameters are foreseen to be not sufficiently accurate \cite{Sayed2019}. In other words, the uncertainty associated with the output of currently employed BEM models is rather large. 
% to arrive at robust BEM models that include an uncertainty estimate associated

Recently, several papers have addressed the uncertainty in BEM model output by performing forward uncertainty propagation or sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019a}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

@@

For the case of a rigid turbine, with a uniform inflow field, the main model uncertainties that arise in the BEM formulation are the following \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included in BEM and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
Next to these model uncertainties we will consider the chord and twist distribution along the blade to be uncertain (geometric uncertainties).

@@

\subsection{NURBS representation}
A NURBS curve $S(r)$ is expressed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(r) = \sum_{i=1}^{n} c_i B_{i,p}(r),
\end{equation}
where $c_i$ is the weight of the $i$-th control point and $B_{i,p}(r)$ is the value of the B-spline corresponding to the $i$-th control point at $x$, with the subscript $p$ denoting its polynomial degree.  In our case, $S(r)$ represent for example the chord as a function of the radial location along the blade. A sequence of non-decreasing parameters $X = \{r_1,r_2,...,r_m\}$, called the \textit{knot vector}, determines the domain in which any control point is active. A control point $c_i$ is said to be active at a location when the corresponding B-spline $B_{i,p}(r)$ is non-zero at $r$. B-splines are recursive in polynomial degree, thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with a B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(r) :=
\begin{cases}
1\quad r_i\leq r < r_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher-order B-splines using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(r) := \frac{r - r_i}{r_{i+p} - r_i}B_{i,p-1}(r) + \frac{r_{i+p+1} - r}{r_{i+p+1} - r_{i+1}}B_{i+1,p-1}(r),\quad p\geq1.
\end{equation}
Note that with increasing $p$ the interval in which a B-spline is non-zero also become larger. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $r$. For example, for a linear curve ($p=1$) only two adjacent control points are active at any location. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. To compute high-order B-splines in \eqref{NURBS_recurrence}, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{basis_examples}, we show linear, quadratic and cubic splines for $r\in[0,1]$ with original knot vector $X = \{0,1/3,2/3,1\}$, padded accordingly with the curve order, for e.g., for linear B-splines, we use the knot vector $X = \{0,0,1/3,2/3,1,1\}$.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 0cm 0 0cm},clip, scale=.75]{basis_example.eps}
\caption{Examples of B-splines with original knot vector $X = \{0,1/3,2/3,1\}$.}
\label{basis_examples}
\end{figure}

In order to parameterize a given curve, e.g. the blade reference chord, we choose a degree $p$ and a number of knots $X =\{r_j\}_{j=1}^n$. The knots can be uniformly spaced or chosen heuristically (for advanced approaches for knot selection see\cite{LI2005791,razdan1999knot}). Next we sample $S_{ref}(r)$ at $X$ and also compute $B_{i,p}(r_j), \text{ for }i, j = 1,2, ..., n$. We then formulate a linear system and solve for the set of control points $\mathbf{c}=\{c_i\}_{i=1}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(r)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with the $j$-th row consisting of values of $n$ B-splines sampled at location $r_j$. Once the control points are obtained, the approximate reference curve $S_n(r)$ follows from \eqref{NURB_curve}. Fig.~\ref{nurbs_approx} shows a parameterization of the chord curve using $n=9$ control points and second-order ($p=2$) B-splines.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Parametrization of chord using 9 control points and second-order B-splines.}
\label{nurbs_approx}
\end{figure}

@@

\subsubsection{Sensitivity analysis results}
In Fig.~\ref{sensitivity_analysis}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output computed using LARS with $M=19$ uncertain parameter chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe the important result that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity compared to the twist variables.

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{}
\label{sensitivity_analysis}
\end{figure}

>>>>

As a consequence of the relatively strong model assumptions at the basis of BEM theory, the results from BEM codes can be subject to significant uncertainties. For example, the effect of sheared inflow \cite{Madsen2012} is not naturally accounted for in the theory and needs to be incorporated via correction terms. Other major model uncertainties in BEM models are the time constant in dynamic stall models, the wake correction factor, the tip loss model parameter, and the lift- and drag-polars used to compute local aerodynamic forces. Especially for increasing turbine sizes, these model parameters are likely not sufficiently accurate \cite{Sayed2019}. 
%In other words, the uncertainty associated with the output of currently employed BEM models is rather large. 
% to arrive at robust BEM models that include an uncertainty estimate associated

Recently, several papers have addressed the uncertainty in BEM model output by performing uncertainty propagation and sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019a}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

@@

For the case of a rigid turbine, with a uniform inflow field, some of the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

\subsection{NURBS representation}
A NURBS curve $S(r)$ is expressed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(r) = \sum_{i=1}^{n} c_i B_{i,p}(r),
\end{equation}
where $c_i$ is the weight of the $i$-th control point and $B_{i,p}(r)$ is the value of the B-spline corresponding to the $i$-th control point at radial location $r$ along the blade, with $p$ denoting the polynomial degree.  In our case, $S(r)$ represents for example the chord length. A sequence of non-decreasing parameters $X = \{r_1,r_2,...,r_m\}$, called the \textit{knot vector}, determines the domain in which any control point is active. A control point $c_i$ is said to be active at a location when the corresponding B-spline $B_{i,p}(r)$ is non-zero at $r$. B-splines are recursive in polynomial degree: quadratic B-splines follow from linear B-splines, cubic from quadratic B-splines etc. Starting with a B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(r) :=
\begin{cases}
1\quad r_i\leq r < r_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
higher-order B-splines follow using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(r) := \frac{r - r_i}{r_{i+p} - r_i}B_{i,p-1}(r) + \frac{r_{i+p+1} - r}{r_{i+p+1} - r_{i+1}}B_{i+1,p-1}(r),\quad p\geq1.
\end{equation}
%Note that with increasing $p$ the interval in which a B-spline is non-zero also become larger. 
%For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $r$. For example, for a linear curve ($p=1$) only two adjacent control points are active at any location. 
Only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}; we use the de Boor's algorithm \cite{deBoor} for efficient construction. 
%To compute high-order B-splines in \eqref{NURBS_recurrence}, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. 
In Fig. \ref{basis_examples}, we show linear, quadratic and cubic splines for $r\in[0,1]$ with original knot vector $X = \{0,1/3,2/3,1\}$, padded accordingly with the curve order, for e.g., for linear B-splines, we use the knot vector $X = \{0,0,1/3,2/3,1,1\}$.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 0cm 0 0cm},clip, scale=.75]{basis_example.eps}
\caption{Examples of B-splines with original knot vector $X = \{0,1/3,2/3,1\}$.}
\label{basis_examples}
\end{figure}

In order to parameterize a given curve, e.g.\ the blade reference chord, we choose a degree $p$ and a number of knots $X =\{r_j\}_{j=1}^n$. The knots can be uniformly spaced or chosen heuristically (for advanced knot selection approaches for knot selection see \cite{Li2005}). Next we sample $S_{ref}(r)$ at $X$ and also compute $B_{i,p}(r_j), \text{ for }i, j = 1,2, ..., n$. We then formulate a linear system and solve for the set of control points $\mathbf{c}=\{c_i\}_{i=1}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(r)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with the $j$-th row consisting of values of $n$ B-splines sampled at location $r_j$. Once the control points are obtained, the approximate reference curve $S_n(r)$ follows from \eqref{NURB_curve}. Fig.~\ref{nurbs_approx} shows a parameterization of the chord curve using $n=9$ control points and second-order ($p=2$) B-splines.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Parametrization of chord using 9 control points and second-order B-splines.}
\label{nurbs_approx}
\end{figure}

@@

\subsubsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade.

\begin{figure}[h!]
\centering
\includegraphics[trim={2cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of power output towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_power}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.5]{SA_axial_force_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of axial force towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_force}
\end{figure}
